Liveness-Based Garbage Collection for Lazy Languages

### 2015-12-01 paper/paper.tex
Fourth submission

\documentclass[preprint, 9pt]{sigplanconf}
\usepackage{etex}
\usepackage{float}
\usepackage{mathrsfs}
%%\usepackage{lineno}
\usepackage{booktabs}
\usepackage{stmaryrd}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
\usepackage{pst-rel-points}
\usepackage{bcprules}
\usepackage{subfigure}
\usepackage[boxed]{algorithm2e}
\usepackage{makecell}
\usepackage{balance}

@@

%% We use the  result of the analysis to annotate  each potential garbage
%% collection  point  in   the  program  with  a   set  of  deterministic
%% finite-state automata (DFA) describing the liveness at that point.
To show the effectiveness of our method, we
implemented a copying collector that uses the results of the liveness
analysis to preserve live 
objects, both evaluated (i.e., in WHNF) and closures.  Our experiments
confirm  that  for  programs  running with  a  liveness-based  garbage
collector,   there  is   a   significant  decrease   in  peak   memory
requirements.   In  addition,   a  sizable  reduction  in   the  number  of
collections  ensures that  in spite  of using  a more  complex garbage
collector,  the execution  time  of  a program  remains comparable  with
the same program running with a reachability-based collector.
\end{abstract}

@@

The variable  \pz\ returned  at $\psi_3$ depends  on \py\  through the
expression $(+~\py~1)$.  $\py$ in turn depends on the closure variable
$\px_{\mathit{\pi}_1}$  through $(\length~\px_{\mathit{\pi}_1})$.   We
denote this  chain of  dependences as  $\lbrack \psi_3\!\!\!:\!\!\!\pz
\leftarrow            (+~\py~1),\,           \py            \leftarrow
(\length~\px_{\mathit{\pi}_1})\rbrack$.  Indeed,  the chains of closures  in the
heap are a  runtime representation of these  dependences.  Since $\pz$
is evaluated  at $\psi_3$  due to  the expression  $\SRETURN~\pz$, the
demand made by the calling context(s)  of $f$ places a demand on $\pz$
which  will  impart a  liveness  to  $\px_{\pi_1}$.  Other  dependence
chains  which result  in  a liveness  for  $\px_{\pi_1}$ are  $\lbrack
\psi_1\!\!\!:\!\!\py   \leftarrow  (\length~\px_{\pi_1})\rbrack$   and
$\lbrack   \psi_2\!\!\!:\!\!\!\pw   \leftarrow  (/~\pu~\py),\,\,   \py
\leftarrow   (\length~\px_{\pi_1})\rbrack$.   The   liveness  analysis
described in this section declares the liveness of $\px_{\pi_1}$ to be
a union of  the liveness arising from these dependence  chains.  To be
safe, a garbage collection during  evaluation of $\py$ at $\psi_1$ has
to use  this liveness to  copy the heap starting from  $\px_{\pi_1}$.  However,
notice that if a garbage collection takes place while evaluating $\pz$
at $\psi_3$, it can safely consider only the liveness arising from the
dependence chain\linebreak  $\lbrack \psi_3\!\!\!:\!\!\!\pz \leftarrow
(+~\py~1),\, \py \leftarrow (\length~\x_{\pi_1})\rbrack$.  The garbage
collection  scheme  described  in Section~\ref{sec:GC-scheme}  uses  a
generalization of this observation  to dynamically select a evaluation
point specific liveness in order to collect more garbage.

@@

\newcommand{\hgt}{2.9cm}
\newcommand{\LGCLine}{blue}
\newcommand{\RGCLine}{red}
\newcommand{\ReachLine}{black}
\newcommand{\UseLine}{lightblue}

@@

As demonstrated by the gap between the \RGCLine\ and the \UseLine, a large number of cells 
which are unused by the program are still copied during RGC. LGC does a much better job of closing this gap 
but still falls short of the precision achieved by LGC in case of eager languages
~\cite{asati14lgc}. 
A major source of inefficiency in LGC is multiple traversals of
already copied heap cells. Since LGC does not mark the heap cells
after the first visit, the same cells can be repeatedly visited with
different liveness states. We have mitigated this problem by
implementing a heuristic which tags \CONS\ cells with the set of
states already visited.

\section{Related work} 
\label{sec:relatedwork}
The important paper
by~\cite{Hirzel} investigates the impact of liveness on the
effectiveness of garbage collection. Their observation is that liveness
can significantly impact garbage collection but only when it is
interprocedural. As far as memory requirement is concerned, our paper
demonstrates this observation.

There have been several attempts to use liveness analysis to improve 
garbage collection for imperative languages.~\cite{khedker07heap}
presents a liveness analysis and uses the results for inserting 
nullifying statements in Java programs. ~\cite{ran.shaham-sas03} is
another work which constructs an automaton modelling heap accesses and
checks temporal properties like liveness against the model.  
However both the approaches are restricted to intraprocedural scope.

In the space of functional languages, efforts towards improving
space efficiency using compile-time  analysis are: rewriting
methods                             such                            as
deforestation~\cite{wadler88deforest,gill93ashort,chitil99deforest},
sharing  analysis   based  reallocation~\cite{jones89compile},  region
based   analysis~\cite{tofte98region},   or   through   insertion   of
compile-time                                                nullifying
statements~\cite{inoue88analysis,lee05static,Hamilton}.            All
compile-time marking approaches rely on an efficient and precise alias
analysis and  cannot provide  significant improvement in  its absence.
An interesting  approach suggested in~\cite{HofmannJ03}  is to
annotate the heap usage of  first-order programs through linear types.
The annotations are then used to serve memory
requests through re-allocation.  However, this
requires the user to write programs in a specific way.

>>>>

\documentclass[9pt]{sigplanconf}
\usepackage{etex}
\usepackage{float}
\usepackage{mathrsfs}
%%\usepackage{lineno}
\usepackage{booktabs}
\usepackage{stmaryrd}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
\usepackage{pst-rel-points}
\usepackage{bcprules}
\usepackage{subfigure}
\usepackage[boxed]{algorithm2e}
\usepackage{makecell}
\usepackage{balance}

@@

%% We use the  result of the analysis to annotate  each potential garbage
%% collection  point  in   the  program  with  a   set  of  deterministic
%% finite-state automata (DFA) describing the liveness at that point.
To  show the  effectiveness of  our method,  we implemented  a copying
collector that uses  the results of the liveness  analysis to preserve
live  objects,  both evaluated  (i.e.,  in  WHNF) and  closures.   Our
experiments confirm  that for  programs running with  a liveness-based
garbage  collector, there  is a  significant decrease  in peak  memory
requirements.   In addition,  a  sizable reduction  in  the number  of
collections  ensures that  in spite  of using  a more  complex garbage
collector, the execution  times of programs running  with liveness and
reachability-based collectors remain comparable.
\end{abstract}

@@

The variable  \pz\ returned  at $\psi_3$ depends  on \py\  through the
expression $(+~\py~1)$.  $\py$ in turn depends on the closure variable
$\px_{\mathit{\pi}_1}$  through $(\length~\px_{\mathit{\pi}_1})$.   We
denote this  chain of  dependences as  $\lbrack \psi_3\!\!\!:\!\!\!\pz
\leftarrow            (+~\py~1),\,           \py            \leftarrow
(\length~\px_{\mathit{\pi}_1})\rbrack$.  Indeed,  the chains of closures  in the
heap are a  runtime representation of these  dependences.  Since $\pz$
is evaluated  at $\psi_3$  due to  the expression  $\SRETURN~\pz$, the
demand made by the calling context(s)  of $f$ places a demand on $\pz$
which  will  impart a  liveness  to  $\px_{\pi_1}$.  Other  dependence
chains  which result  in  a liveness  for  $\px_{\pi_1}$ are  $\lbrack
\psi_1\!\!\!:\!\!\py   \leftarrow  (\length~\px_{\pi_1})\rbrack$   and
$\lbrack   \psi_2\!\!\!:\!\!\!\pw   \leftarrow  (/~\pu~\py),\,\,   \py
\leftarrow   (\length~\px_{\pi_1})\rbrack$.   The   liveness  analysis
described in this section declares the liveness of $\px_{\pi_1}$ to be
a union of  the liveness arising from these dependence  chains.  To be
safe, a garbage collection during  evaluation of $\py$ at $\psi_1$ has
to use  this liveness to  copy the heap starting from  $\px_{\pi_1}$.  However,
notice that if a garbage collection takes place while evaluating $\pz$
at $\psi_3$, it can safely consider only the liveness arising from the
dependence chain\linebreak  $\lbrack \psi_3\!\!\!:\!\!\!\pz \leftarrow
(+~\py~1),\, \py \leftarrow (\length~\px_{\pi_1})\rbrack$.  The garbage
collection  scheme  described  in Section~\ref{sec:GC-scheme}  uses  a
generalization of this observation  to dynamically select a evaluation
point specific liveness in order to collect more garbage.

@@

\newcommand{\hgt}{2.9cm}
\newcommand{\LGCLine}{blue}
\newcommand{\RGCLine}{red}
\newcommand{\ReachLine}{black}
\newcommand{\UseLine}{light-blue}

@@

As  demonstrated by  the gap  between the  \RGCLine\ and  the \UseLine\
lines, a  large number of  cells which are  unused by the  program are
still copied  during RGC. LGC does  a much better job  of closing this
gap but still falls short of the  precision achieved by LGC in case of
eager languages ~\cite{asati14lgc}.  A major source of inefficiency in
LGC is  multiple traversals  of already copied  heap cells.  Since LGC
does not mark the heap cells after the first visit, the same cells can
be  repeatedly  visited  with   different  liveness  states.  We  have
mitigated  this  problem  by   implementing  a  heuristic  which  tags
\CONS\ cells with the set of states already visited.

\section{Related work} 
\label{sec:relatedwork}
The impact of  liveness on the effectiveness of  garbage collection is
investigated in~\cite{Hirzel}.   The observation is that  liveness can
significantly  impact   garbage  collection,  but  only   when  it  is
interprocedural.  As far as memory requirement is concerned, our paper
demonstrates this observation.

There have been several attempts to use liveness analysis to improve 
garbage collection for imperative languages.~\cite{khedker07heap}
presents a liveness analysis and uses the results for inserting 
nullifying statements in Java programs. In ~\cite{ran.shaham-sas03},
an automaton modelling heap accesses is created, and
temporal properties like liveness are checked  against the model.  
However both the approaches are restricted to intraprocedural scope.

In the space of functional  languages, there are: rewriting methods such as
deforestation~\cite{wadler88deforest,gill93ashort,chitil99deforest},
sharing  analysis   based  reallocation~\cite{jones89compile},  region
based   analysis~\cite{tofte98region},   and    insertion   of
compile-time                                                nullifying
statements~\cite{inoue88analysis,lee05static,Hamilton}.            All
compile-time marking approaches rely on an efficient and precise alias
analysis and  cannot provide  significant improvement in  its absence.
An interesting approach suggested  in~\cite{HofmannJ03} is to annotate
the  heap usage  of first-order  programs through  linear types.   The
annotations   are  then   used  to   serve  memory   requests  through
re-allocation.  However, this requires the user to write programs in a
specific way.
